\documentclass{article}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{natbib}

\input{math-macros.tex}

\title{Experimental Roadmap for Validating Fenchel--Young Surrogates in Kidney Exchange}
\author{Weikang}
\date{}

\begin{document}
\maketitle

\section{Motivation}

The purpose of the experiments is not only to compare decision-focused learning
(DFL) against a two-stage predict-then-optimize pipeline, but also to establish
whether the Fenchel--Young (FY) perturbed-optimizer loss
\citep{berthet2020learning} is a meaningful surrogate for the true decision
loss in our kidney exchange setting.

This distinction matters because the realistic synthetic benchmark used in our
main experiments has a complex reward mechanism. The ground-truth edge reward is
not a simple linear function of utility, but a nonlinear product of transplant
success probability, QALY, scarcity priority, graph-structural variables, and
deterministic edge-level noise. Therefore, a direct comparison between DFL and a
two-stage baseline on this benchmark alone cannot isolate the behavior of the
surrogate loss. If DFL improves the final matching score, the improvement may be
due to the surrogate, the choice of perturbation scale, the warm start, or model
capacity. If DFL does not improve, the failure may instead be caused by model
misspecification or an overly complex label-generation process.

We therefore separate the experimental study into several layers, moving from a
clean surrogate-validity test to the full realistic synthetic benchmark.

\section{Experimental Roadmap}

\subsection{Step 1: Surrogate Validity Under a Clean Linear Label}

The first experiment should validate whether the FY loss is directionally aligned
with the true regret in a controlled setting. We use a label-generation mechanism
that is exactly representable by the prediction model:
\[
    w^{\mathrm{true}}_e = a u_e + b c_e,
\]
where \(u_e\) is the edge utility and \(c_e\) is the recipient cPRA associated
with edge \(e\). The corresponding prediction model is
\[
    \hat w_e(\theta) = \theta_1 u_e + \theta_2 c_e.
\]

This setting removes model misspecification. If the FY surrogate is appropriate,
its minimizer should lie in or near the low-regret region of the true decision
loss landscape.

The main diagnostic objects are:
\begin{itemize}
    \item the true regret landscape \(R(\theta_1,\theta_2)\);
    \item the FY surrogate landscape \(L_{\mathrm{FY}}(\theta_1,\theta_2)\);
    \item the two-stage MSE solution \(\theta^\star_{\mathrm{2s}}\);
    \item the true oracle set
    \[
        \Theta^\star = \arg\min_{\theta} R(\theta);
    \]
    \item the FY minimizer
    \[
        \theta^\star_{\mathrm{FY}}(\epsilon)
        =
        \arg\min_{\theta} L_{\mathrm{FY},\epsilon}(\theta).
    \]
\end{itemize}

The primary metric should be regret gap rather than parameter distance:
\[
    \Delta R(\epsilon)
    =
    R(\theta^\star_{\mathrm{FY}}(\epsilon))
    -
    \min_{\theta} R(\theta).
\]
This is important because the KEP solver is piecewise constant, so the true
regret landscape often has an optimal plateau rather than a unique isolated
minimizer. In such cases, parameter-space distance to one selected oracle point
is not well-defined, whereas \(\Delta R\) remains meaningful.

This step answers the question:

\begin{quote}
Does the FY surrogate point toward the true low-regret region when the predictive
model is correctly specified?
\end{quote}

\subsection{Step 2: Controlled Model Misspecification}

The second experiment should introduce a limited amount of misspecification.
For example, the true reward may include an additional feature or interaction:
\[
    w^{\mathrm{true}}_e
    =
    a u_e + b c_e + d\,\mathrm{age}^{\mathrm{donor}}_e,
\]
or
\[
    w^{\mathrm{true}}_e
    =
    a u_e + b c_e + d\,u_e c_e.
\]
The prediction model can then be kept deliberately simpler:
\[
    \hat w_e(\theta) = \theta_1 u_e + \theta_2 c_e.
\]

This setting tests whether FY optimization is robust when the model cannot
perfectly recover the edge rewards. The question is no longer whether FY can
recover the true label function, but whether it can still produce better
decisions than an MSE-trained predictor.

The same landscape diagnostics should be reported:
\[
    R(\theta), \quad
    L_{\mathrm{FY},\epsilon}(\theta), \quad
    R(\theta^\star_{\mathrm{2s}}), \quad
    R(\theta^\star_{\mathrm{FY}}), \quad
    \Delta R.
\]

This step answers the question:

\begin{quote}
When the predictive model is misspecified, does the FY surrogate still move the
parameters toward lower true regret than the two-stage MSE solution?
\end{quote}

\subsection{Step 3: Training-Level DFL vs Two-Stage Comparison}

After the landscape-level surrogate diagnostics, we should run actual training
experiments on the same controlled settings. Both methods should use the same
train/validation/test split, the same model class, and the same KEP solver.

The two-stage baseline first minimizes prediction error:
\[
    \min_{\theta}
    \sum_{e}
    \left(\hat w_e(\theta)-w^{\mathrm{true}}_e\right)^2,
\]
then solves the KEP using the predicted weights \(\hat \bw(\theta)\).

The DFL model starts either from the two-stage checkpoint or from random
initialization and then updates the predictor using the FY gradient estimate:
\[
    \nabla_{\hat \bw} L_{\mathrm{FY}}
    \approx
    \bar y_{\epsilon}(\hat \bw) - y^\star,
\]
where
\[
    \bar y_{\epsilon}(\hat \bw)
    =
    \frac{1}{M}
    \sum_{m=1}^{M}
    S(\hat \bw+\epsilon z^{(m)}),
    \qquad
    z^{(m)}\sim \mathcal{N}(0,I).
\]

The primary evaluation metric should be final achieved ground-truth score, or
equivalently average regret:
\[
    R(\theta)
    =
    \frac{1}{N}
    \sum_{i=1}^{N}
    \left[
        (\bw^i)^\top y^\star(\bw^i)
        -
        (\bw^i)^\top S(\hat \bw^i(\theta))
    \right].
\]

This step answers the question:

\begin{quote}
Does the landscape-level alignment between FY and true regret translate into a
training-level improvement over the two-stage baseline?
\end{quote}

\subsection{Step 4: Realistic Synthetic Benchmark}

The final main benchmark should use the currently processed dataset, now named
\texttt{dataset/processed/realistic\_synthetic\_benchmark}. This benchmark uses
the nonlinear medical reward:
\[
    w^{\mathrm{true}}_e
    =
    E[\mathrm{transplant}]_e
    \cdot
    \mathrm{QALY}_e
    \cdot
    \mathrm{priority}_e
    \cdot
    (1+\xi_e).
\]
This label incorporates utility, recipient cPRA, donor and recipient age, blood
type context, arc failure probability, vertex failure probability, scarcity,
donor flexibility, reciprocal edges, and deterministic edge-level noise.

This benchmark should be described as a realistic synthetic benchmark rather
than a clean surrogate-identification experiment. Its role is to test whether
the DFL pipeline improves exchange-plan quality in a richer synthetic
environment.

The interpretation should be careful:
\begin{itemize}
    \item If DFL improves over two-stage, this provides empirical evidence that
    the FY-based pipeline can help in a complex KEP setting.
    \item If DFL does not improve, this does not by itself prove that the FY
    surrogate is invalid; the failure may come from model misspecification,
    insufficient features, perturbation scale, or optimization instability.
\end{itemize}

This step answers the question:

\begin{quote}
Under a medically motivated synthetic reward model, does FY-based DFL improve
the final kidney exchange objective compared with the two-stage pipeline?
\end{quote}

\subsection{Step 5: Sensitivity and Ablation Studies}

The last layer should quantify which design choices affect the result. The most
important ablations are:
\begin{itemize}
    \item perturbation scale \(\epsilon\);
    \item number of perturbation samples \(M\);
    \item warm start from the two-stage solution versus random initialization;
    \item feature set: utility only, utility+cPRA, small linear features,
    failure context features, and full features;
    \item label noise on versus off;
    \item solver formulation, such as CF, PIEF, and hybrid.
\end{itemize}

For the perturbation scale, it is important to distinguish absolute and relative
noise. Since the KEP solver is an argmax, its output is invariant to positive
rescaling of all weights. Therefore the effective smoothing strength is governed
by the relative scale of the perturbation:
\[
    S(\hat \bw+\epsilon z)
    =
    S\left(\frac{\hat \bw}{\epsilon}+z\right).
\]
Thus, \(\epsilon\) should be interpreted relative to the scale of
\(\hat \bw\). In the landscape plots, \(\epsilon\) may be calibrated relative to
the MSE weight scale, while in end-to-end training it may be used as an absolute
noise standard deviation. This difference should be reported explicitly.

\section{Recommended Presentation Order}

The final paper should present the experiments in the following order:
\begin{enumerate}
    \item clean linear surrogate-validity benchmark;
    \item controlled misspecification benchmark;
    \item training-level DFL versus two-stage comparison on controlled labels;
    \item realistic synthetic benchmark;
    \item epsilon, sample-size, feature-set, and initialization ablations.
\end{enumerate}

This ordering separates mechanism validation from final empirical performance.
It allows the paper to make precise claims: first, whether the FY surrogate is
aligned with true regret under controlled conditions; second, whether this
alignment survives moderate misspecification; and finally, whether it yields
practical improvements on the realistic synthetic KEP benchmark.

\bibliographystyle{plainnat}
\bibliography{ref}

\end{document}
