\section{Experiments and Diagnostics}

We now test the central hypothesis from the previous section: in
decomposable packing-style problems, prediction errors may change the identity
of the selected solution without substantially changing its true objective
value. We focus on kidney exchange, where a feasible decision is a set of
vertex-disjoint cycles and chains. All kidney-exchange experiments use the
Step2b-d8 label regime and compare a two-stage model selected by validation MSE
with an SPO+ model selected by validation SPO+ loss. Unless otherwise stated,
we use maximum cycle length three and maximum chain length four.

For each graph and method, we compute the best predicted solution, denoted
rank~1. We then add a no-good cut excluding this solution and re-solve to obtain
the second-best distinct predicted solution, denoted rank~2. Both solutions are
evaluated using the true edge weights and compared against the oracle solution.
The main quantity of interest is the normalized oracle gap:
\[
    \mathrm{gap}(y)
    =
    \frac{\mathrm{obj}(y^\star)-\mathrm{obj}(y)}
    {|\mathrm{obj}(y^\star)|},
\]
where $y^\star$ is the oracle solution and $\mathrm{obj}(\cdot)$ is the true
objective value. A rank-2 solution within $5\%$ of the oracle indicates that the
second-best predicted solution is still a high-quality alternative.

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.34\linewidth}
                >{\raggedright\arraybackslash}p{0.58\linewidth}@{}}
\toprule
Item & Setting \\
\midrule
Regime & Step2b-d8 \\
Train seeds & 9 selected training seeds \\
Held-out graphs & 400 \\
Graph-seed cases & 3600 per method and rank \\
Methods & \texttt{2stage\_val\_mse}; \texttt{spoplus\_val\_spoplus\_loss} \\
Cycle and chain limits & Max cycle 3 unless stated; max chain 4 \\
Rank-2 computation & No-good cut excluding the rank-1 predicted solution \\
Metrics & Normalized oracle gap, within-5\% rate, exact oracle rate, edge Jaccard \\
\bottomrule
\end{tabular}
\caption{Compact experimental setup for the KEP diagnostics.}
\label{tab:experiment-setup}
\end{table}

\subsection{Close Second-Best Solutions in KEP}

\paragraph{Question.}
Do KEP instances contain close alternative solutions near the top of the
predicted ranking?

\paragraph{Setup.}
We evaluate 9 selected training seeds and 400 held-out graphs, giving 3600
graph-seed cases for each method and rank. For each case, we compute rank~1 and
rank~2 solutions for 2stage and SPO+ and evaluate both under the true weights.

\paragraph{Results.}
Table~\ref{tab:second-best-summary} summarizes the main rank-1/rank-2
comparison. SPO+ improves the best predicted solution over 2stage, reducing the
mean normalized gap from $6.09\%$ to $4.94\%$. However, the rank-2 solutions do
not collapse in quality. The 2stage rank-2 solution has mean normalized gap
$7.04\%$, and the SPO+ rank-2 solution has mean normalized gap $6.32\%$.
Moreover, $45.19\%$ of 2stage rank-2 solutions and $47.97\%$ of SPO+ rank-2
solutions are within $5\%$ of the oracle.

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}llrrrr@{}}
\toprule
Method & Rank & Mean & Median & Exact & Within 5\% \\
\midrule
2stage & 1 & 6.09\% & 4.38\% & 17.72\% & 53.14\% \\
2stage & 2 & 7.04\% & 5.75\% & 8.14\% & 45.19\% \\
SPO+ & 1 & 4.94\% & 3.46\% & 23.44\% & 59.28\% \\
SPO+ & 2 & 6.32\% & 5.28\% & 7.19\% & 47.97\% \\
\bottomrule
\end{tabular}
\caption{Best and second-best KEP solutions under max-cycle 3. Rank~2 is
obtained by excluding the rank-1 solution with a no-good cut.}
\label{tab:second-best-summary}
\end{table}

\paragraph{Interpretation.}
These results directly support the close-substitute hypothesis. In many KEP
instances, excluding the best predicted solution still leaves a second predicted
solution whose true objective value is close to the oracle. Thus, changing
solution identity is not always equivalent to incurring large decision regret.
This helps explain why the average improvement of SPO+ over 2stage is moderate:
when 2stage already selects a near-oracle alternative, there is little remaining
objective gap for a decision-focused method to close.

\subsection{Cycle-Length Sensitivity}

\paragraph{Question.}
Is the closeness of rank-2 solutions merely an artifact of restricting cycles
to length three?

\paragraph{Setup.}
We keep the same trained models, graph split, and selected seeds, and only
change the KEP feasible set by allowing maximum cycle length $K \in
\{3,4,5\}$. The maximum chain length remains fixed at four. For each $K$, we
recompute the oracle, rank~1, and rank~2 solutions.

\paragraph{Results.}
Table~\ref{tab:cycle-summary} reports the rank-2 results across cycle lengths.
The rank-2 gap does not increase as longer cycles are allowed. For 2stage, the
rank-2 mean normalized gap is essentially stable: $7.04\%$ for $K=3$,
$6.93\%$ for $K=4$, and $6.88\%$ for $K=5$. For SPO+, the rank-2 gap decreases
from $6.32\%$ to $5.61\%$. The within-$5\%$ rate also increases for both
methods.

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}llrrr@{}}
\toprule
$K$ & Method & Mean gap & Median gap & Within 5\% \\
\midrule
3 & 2stage & 7.04\% & 5.75\% & 45.19\% \\
4 & 2stage & 6.93\% & 5.53\% & 47.50\% \\
5 & 2stage & 6.88\% & 5.20\% & 48.94\% \\
3 & SPO+ & 6.32\% & 5.28\% & 47.97\% \\
4 & SPO+ & 6.20\% & 5.03\% & 49.78\% \\
5 & SPO+ & 5.61\% & 4.39\% & 54.83\% \\
\bottomrule
\end{tabular}
\caption{Cycle-length sensitivity. Increasing the maximum cycle length from 3
to 4 or 5 does not make rank-2 solutions systematically farther from the
oracle.}
\label{tab:cycle-summary}
\end{table}

\paragraph{Interpretation.}
The max-cycle-3 observation is robust to longer cycles. Although enlarging the
feasible set changes the oracle and many selected solutions, the second-best
solutions remain close to the oracle in aggregate. Therefore, the presence of
close rank-2 alternatives is not simply a consequence of the max-cycle-3
restriction. We treat oracle improvement under larger feasible sets as a sanity
check rather than a main finding, since it is expected when more feasible
solutions are allowed.

\subsection{Arc-Density Sensitivity}

\paragraph{Question.}
Does the close-rank-2 behavior persist when the graph density changes?

\paragraph{Setup.}
We perform a selected-case density sensitivity study on three
mechanism-representative graphs: G-696, a close-alternatives case; G-392, an
SPO+ region-correction case; and G-1560, a rank-2 promotion case. We keep
max-cycle 3 and max-chain 4 fixed and generate four arc-density variants:
adding $25\%$ arcs, adding 25 arcs, removing 25 arcs, and removing $25\%$ arcs.
The fixed-count removal follows the advisor-requested decrease condition, while
the percentage removal is treated as a robustness variant. Because arc
perturbation is random, we repeat the generation with perturbation seeds
0--4. This gives 15 graph-perturbation cases per variant and method.

\paragraph{Results.}
Table~\ref{tab:density-summary} reports the density sensitivity graph by graph.
This is important because the three graphs were selected to represent different
mechanisms, so pooling them into one mean would hide the case-level behavior. For
the add-arc and remove-arc rows, each cell reports the range of mean rank-2 gaps
across the fixed-count and percentage variants, averaged over perturbation seeds
0--4 within each variant.

\begin{table*}[t]
\centering
\small
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}llllll@{}}
\toprule
Graph & Mechanism & Variant summary & 2stage rank-2 gap & SPO+ rank-2 gap & Preserved? \\
\midrule
G-696 & Close alternatives & Original & 1.02\% & 3.75\% & Yes \\
G-696 & Close alternatives & Add arcs & 2.40--4.69\% & 3.49--4.31\% & Yes \\
G-696 & Close alternatives & Remove arcs & 2.25--3.58\% & 1.75--5.83\% & Mixed \\
\midrule
G-392 & SPO+ correction & Original & 31.94\% & 2.53\% & Yes \\
G-392 & SPO+ correction & Add arcs & 5.45--10.01\% & 5.37--8.74\% & Yes \\
G-392 & SPO+ correction & Remove arcs & 6.87--12.96\% & 7.20--12.60\% & Mixed \\
\midrule
G-1560 & Rank-2 promotion & Original & 1.20\% & 29.00\% & Yes \\
G-1560 & Rank-2 promotion & Add arcs & 7.35--7.87\% & 4.83--4.90\% & Mixed \\
G-1560 & Rank-2 promotion & Remove arcs & 4.12--10.48\% & 5.99--15.88\% & Mixed \\
\bottomrule
\end{tabular}
\caption{Per-graph arc-density sensitivity on three selected mechanism graphs.
Rank-2 gaps are normalized oracle gaps. Add/remove rows summarize the two
corresponding arc-density variants as ranges over their seed-averaged gaps. These
are selected-case mechanism results, not population-level statistics.}
\label{tab:density-summary}
\end{table*}

The per-graph view clarifies what the pooled average obscures. G-696 remains a
close-alternative case under most perturbations. G-392 starts as a strong
two-stage failure corrected by SPO+, but density changes soften the original
failure mode and reduce the separation between methods. G-1560 preserves the
original rank-2 promotion relation only in the unperturbed graph; after arc
changes, the exact promotion mechanism becomes mixed even though rank-2 gaps
remain informative.

The oracle changes in the expected direction: add-arc variants increase the
oracle objective in all selected graph-perturbation cases, and remove-arc
variants decrease it in all selected graph-perturbation cases. More importantly,
the added arcs are actively used by the oracle and by the learned rank-1/rank-2
solutions. Thus, the density perturbations genuinely reshape the feasible
landscape rather than adding irrelevant arcs.

\paragraph{Interpretation.}
The density study is more mechanism-sensitive than the cycle-length study.
Adding arcs tends to create additional high-quality alternatives in these
selected graphs. Removal is more dependent on which arcs are removed: some
seeds mildly worsen rank-2 gaps, while others improve them. We therefore use
the density results as a mechanism stress test rather than as a full-population
estimate. The key conclusion is that close second-best behavior is not driven by
a single perturbation seed, but the exact mechanism varies by graph.

\FloatBarrier

\subsection{Graph-Level Mechanism Case Studies}

\paragraph{Question.}
What mechanisms are hidden by the aggregate density averages?

\paragraph{Setup.}
The three density-sensitivity graphs were selected because they represent
different mechanisms in the original max-cycle-3 analysis. We therefore
interpret density sensitivity graph by graph, using the multi-seed robustness
runs to avoid over-interpreting a single perturbation seed.

\paragraph{Results and interpretation.}
Table~\ref{tab:case-mechanisms} summarizes the mechanism-level interpretation.
G-696 mostly preserves the close-alternatives behavior: both methods keep
rank-2 gaps near or below $5\%$ under most add/remove perturbations. G-392
mostly preserves the SPO+ region-correction behavior under moderate
perturbations, but density changes the original 2stage failure mode and makes
the SPO+ advantage less distinguishable under large percentage removal. G-1560
is the most fragile: the exact relation ``2stage rank~2 becomes SPO+ rank~1''
is stable in the original graph but becomes seed-sensitive after arc
perturbations.

\begin{table*}[t]
\centering
\small
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.11\linewidth}
                >{\raggedright\arraybackslash}p{0.22\linewidth}
                >{\raggedright\arraybackslash}p{0.30\linewidth}
                >{\raggedright\arraybackslash}p{0.29\linewidth}@{}}
\toprule
Graph & Original mechanism & Density effect & Interpretation \\
\midrule
G-696 &
Close alternatives. Original rank-2 gaps are $1.02\%$ for 2stage and
$3.75\%$ for SPO+. &
Add-arc and fixed-removal variants mostly keep rank-2 gaps below $5\%$.
The larger removal variant is mixed for SPO+. &
The close-alternative mechanism is mostly preserved. Density perturbation does
not improve the graph so much as preserve an already-close rank-2 structure. \\
\midrule
G-392 &
SPO+ region correction. Original 2stage rank-2 gap is $31.94\%$, while
SPO+ rank-2 is $2.53\%$ and SPO+ rank-1 is oracle-level. &
Perturbations improve the original 2stage failure mode but make SPO+ rank-2
less close than in the original graph. SPO+ remains closer under moderate
perturbations. &
The SPO+ correction mechanism is mostly preserved, but large percentage removal
weakens the distinction between 2stage and SPO+. \\
\midrule
G-1560 &
Rank-2 promotion. In the original graph, the near-oracle 2stage rank-2 solution
is exactly SPO+ rank-1. &
After density perturbation, exact promotion becomes seed-sensitive. Rank-2 gaps
remain informative, but the same edge-set promotion is no longer stable. &
The graph still has useful second-best structure, but the exact promotion
mechanism is fragile under arc perturbation. \\
\bottomrule
\end{tabular}
\caption{Per-graph mechanism interpretation for density sensitivity. The
aggregate averages summarize perturb-seed robustness; this table explains which
mechanisms are preserved or changed.}
\label{tab:case-mechanisms}
\end{table*}

\paragraph{Summary.}
The experiments support a consistent picture. In KEP, rank-2 solutions are often
close to the oracle, and this remains true under longer cycle lengths. Density
perturbations change the feasible landscape more sharply, but the selected
case studies show that close alternatives can persist, while the exact mechanism
behind a method's improvement can be graph-dependent. These empirical patterns
are consistent with the proposed property: close substitute solutions in a
decomposable packing-style feasible set can limit the average advantage of
decision-focused learning over a strong two-stage model.
