\section{Introduction}

Predict-then-optimize pipelines are often evaluated by prediction error even
though the final objective is decision quality. Decision-focused learning (DFL)
addresses this mismatch by training predictors through an optimization-aware
surrogate such as SPO+ \citep{elmachtoub2022smart}. In shortest-path benchmarks,
this idea can produce large gains over a two-stage mean-squared-error baseline.
In our kidney-exchange experiments, however, the gains are more modest.

This paper asks a targeted question: when is DFL useful for graph combinatorial
optimization? Our starting point is that selected-solution identity is not
enough. A model can select a different feasible solution from the oracle and
still achieve nearly the same objective value. This can happen in decomposable
packing-style problems, where many feasible solutions are assembled from mostly
independent components and close substitutes are common.

We therefore analyze prediction errors at the decision level. Rather than asking
only which edge weights are predicted poorly, we ask whether the badly predicted
edges participate in the symmetric difference between selected and oracle
solutions, and whether the alternative solution is truly worse.

\paragraph{Contributions.}
We make four contributions.
\begin{itemize}
    \item We identify close substitute solutions as a mechanism explaining why
    DFL can have limited gains in KEP-like packing problems, even when it
    changes the selected solution.
    \item We introduce a best/second-best diagnostic based on no-good cuts,
    measuring whether alternative predicted solutions remain close to the oracle
    under the true objective.
    \item We validate the mechanism empirically through KEP rank-2 analysis,
    cycle-length sensitivity, density sensitivity, and graph-level case studies.
    \item We provide randomized toy experiments across KEP-like set-packing,
    stable-set, and shortest-path families, showing that identity mismatch and
    decision regret decouple when close substitutes exist.
\end{itemize}
