\section{Mechanism: Close Substitute Solutions}

Our central hypothesis is that DFL is most useful when prediction errors change
the objective value of the selected solution, not merely its identity. This
distinction is important in graph combinatorial optimization because two
different feasible solutions can have very similar true objective values.

\paragraph{Property X.}
We use Property X to denote close substitute solutions in a decomposable
packing-style feasible set. Informally, a problem has this property when a
feasible decision is assembled from several mostly independent components, and
one component can often be replaced by another feasible component with only a
small change in true objective value. In such problems, prediction errors may
change the selected solution identity without creating large decision regret.

\paragraph{Decision identity vs. decision value.}
Let $y^\star$ be the oracle solution under the true objective and let
$y_{\mathrm{2stage}}$ be the solution selected by a two-stage predictor. The fact
that $y_{\mathrm{2stage}} \neq y^\star$ is not, by itself, evidence of a harmful
decision error. The relevant quantity is the normalized oracle gap,
\[
    \mathrm{gap}(y)
    =
    \frac{\mathrm{obj}(y^\star)-\mathrm{obj}(y)}
    {|\mathrm{obj}(y^\star)|},
\]
for reward-maximization problems. A different selected solution with small
$\mathrm{gap}(y)$ is a close substitute, not a catastrophic decision failure.

\paragraph{A simple implication.}
This observation creates an immediate ceiling on possible improvement. If
\[
    \mathrm{gap}(y_{\mathrm{2stage}}) \le \epsilon,
\]
then the maximum possible normalized improvement of any method over two-stage
on this instance is at most $\epsilon$. Indeed, even an oracle-perfect method
can reduce the gap only from $\mathrm{gap}(y_{\mathrm{2stage}})$ to zero. Thus,
when two-stage already selects an $\epsilon$-near-oracle alternative, a
decision-focused method has little remaining objective gap to close on that
instance.

\paragraph{Improvement ceiling.}
This ceiling does not imply that DFL is unnecessary. It says that the benefit of
DFL is instance-dependent. DFL can matter when prediction errors move the
optimizer to a much worse solution. But in instances with many close
substitutes, improving the prediction objective or even recovering the exact
oracle identity may yield only limited normalized-regret improvement. This is
the mechanism we test empirically: KEP may have many non-oracle solutions that
are still near-optimal under the true objective.

\paragraph{Why KEP fits.}
Kidney-exchange decisions are packings of vertex-disjoint cycles and chains.
This structure is decomposable: a feasible solution can often exchange one
cycle or chain for another while preserving feasibility elsewhere in the graph.
When several cycles or chains have similar true value, different packings can
be close substitutes. Therefore, a two-stage model may select a different set of
cycles/chains from the oracle while still achieving a near-oracle objective
value.

\paragraph{Why shortest path differs.}
Shortest-path decisions have a more coupled serial structure. Selecting one
wrong intermediate segment can redirect the entire path, and the resulting path
need not be a local close substitute for the oracle path. Thus, a similar
prediction-order error can create much larger regret in shortest path than in a
decomposable packing problem. This contrast explains why shortest-path
benchmarks can show large DFL gains while KEP shows more moderate gains.
